A model-based method for measuring stability and change in personal culture with three-wave panel data

### 2020-12-17 ms.Rmd

---
title: "A model-based method for measuring stability and change in personal culture with three-wave panel data"
author:
  - Stephen Vaisey, Duke University
  - Kevin Kiley, Duke University
date: "11/22/2020"
output: 
  bookdown::pdf_book:
    number_sections: true
    toc: false
bibliography: ["panel-change-sem.bib"]
---

@@

Kiley and Vaisey [-@kiley2020, hereafter KV] distinguish between two main models of cultural change implied in the literature: the *active updating model* and the *settled dispositions model*. The assumptions of each model are represented in Figure 1.

@@

Estimating this model on 183 variables from the 2006-2014 GSS panels, KV find that a high proportion of outcome variables show no evidence of active updating. Although they point out several exceptions (e.g., some questions about gay rights, where attitudes do appear to be changing over the study period) they conclude that the settled dispositions model is a better default for thinking about public opinion among adults.

@@

If $Y_t$ were always measured perfectly, we could simply compare the penalized likelihoods (e.g., BICs) of the two models to see which is more likely to be the true model given the data [@bollen2014]. Unfortunately, most GSS items likely contain some measurement error. Any measurement bias--the systematic tendency for an item to over- or underestimate a person's "true" value--will induce error correlations between waves that have no basis in the causal process. Thus $U_i$ in Equation (2) represents both settled dispositions and respondent-specific measurement bias.

@@

```{r, out.width="300px", fig.cap="General Model", fig.align='center'}
knitr::include_graphics("figure2.jpg")

@@

@kiley2020 also compare estimates using Equation (1) with and without constraints to adjudicate between the active updating and settled dispositions accounts. The approach we outline here is essentially identical in spirit but a bit different in the details. The main difference is that KV compare models based solely on how well they "predict" the Wave 3 response. Using structural equation models allows comparing the fit of the model to all the observed data, not just the final response.

@@

We use the same 183 variables and three panels of the General Social Survey as @kiley2020. Rather than pool the panels, we estimate models separately on each one. Most variables appear in all three panels but 6 are measured in only one panel and one is measured in two panels. This gives us 536 total three-wave datasets.

@@

@@

ggplot(win_table, aes(y = verdict, x = wincount)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  scale_y_discrete(labels = c("Active Updating", "Inconclusive", "Settled Dispositions")) +
  scale_x_continuous(limits = c(0, 450)) +
  labs(y = "",
       x = "Number of variable-panels")

@@

We have presented a model-based approach to estimating...

>>>>

---
title: "A model-based method for measuring stability and change in personal culture with three-wave panel data"
author:
  - Stephen Vaisey, Duke University
  - Kevin Kiley, Duke University
date: "12/16/2020"
output: 
  bookdown::pdf_book:
    number_sections: true
    toc: false
bibliography: ["panel-change-sem.bib"]
---

@@

Kiley and Vaisey [-@kiley2020, hereafter KV] distinguish between two main models of cultural change implied in the literature: the *active updating model* and the *settled dispositions model*. The assumptions of each model are represented in Figure 1. The active updating model assumes that shocks affecting a person's response to a question at one point in time persist to the next time period, while the settled dispositions model assumes that shocks do not persist.

@@

Estimating this model on 183 variables from the 2006-2014 GSS panels, KV find that about 40 percent of the variables show no evidence of active updating, and most of the remainder show only weak evidence of active updating. Although they point out several exceptions (e.g., some questions about gay rights, where attitudes do appear to be changing over the study period) they conclude that the settled dispositions model is a better default for thinking about public opinion among adults.

@@

If $Y_t$ were always measured perfectly, we could simply compare the penalized likelihoods (e.g., Bayesian Information Criteria) of the two models to see which is more likely to be the true model given the data [@bollen2014]. Unfortunately, most GSS items likely contain some measurement error. Any measurement bias--the systematic tendency for an item to over- or underestimate a person's "true" value--will induce error correlations between waves that have no basis in the causal process. Thus $U_i$ in Equation (2) represents both settled dispositions and respondent-specific measurement bias.

@@

```{r, out.width="400px", fig.cap="General Model", fig.align='center'}
knitr::include_graphics("figure2.pdf")

@@

@kiley2020 also compare estimates using Equation (1) with and without constraints to adjudicate between the active updating and settled dispositions accounts. The approach we outline here is essentially identical in spirit but a bit different in the details. The main difference is that KV compare models based solely on how well they "predict" the Wave 3 response. Using structural equation models allows comparing the fit of the model to all the observed data, not just the final response. For example, the KV model has no leverage in cases where respondents give the same response in waves 1 and 2, but a different response in wave 3, while the model presented here does.

To oversimplify slightly, the models presented here essentially adjudicate whether people are equally likely to deviate from their baseline response in all waves, which would be evidence of the settled dispositions model, or whether waves 1 and 3 are more likely than wave 2 to be deviations, which would be evidence of persisting changes and active updating.

@@

We use the same 183 variables and three panels of the General Social Survey as @kiley2020. Rather than pool the panels, we estimate models separately on each one. Most variables appear in all three panels but 6 are measured in only one panel and one is measured in two panels. This gives us 536 total three-wave data sets.

@@

The approach is quite conservative, favoring the active updating model over the settled dispositions model when there is meaningful evidence that some amount of people in the sample are making durable changes in opinion or behavior. This should not be taken as evidence that *many* people in the population are making durable changes, only that there is any evidence of durable change at all.

@@

ggplot(win_table, aes(y = verdict, x = wincount)) +
  geom_bar(stat = "identity", color = "black", fill = "gray") +
  theme_minimal() +
  scale_y_discrete(labels = c("Active Updating", "Inconclusive", "Settled Dispositions")) +
  scale_x_continuous(limits = c(0, 450)) +
  labs(y = "",
       x = "Number of variable-panels")

@@

Building on KV, we have presented a model-based approach to estimating whether the pattern of attitude and behavior change in the population should be best thought of as following a *settled dispositions* model, where people have stable baselines and change is temporary, or an *active updating* model, where changes tend to persist. Using a set of structural equation models that formalize these theoretical processes under a variety of assumptions, we compared the best-fitting settled dispositions model to the best-fitting active updating model for each variable in each panel, a total of 536 variable-panel pairs. The overwhelming majority of variable-panels prefer the settled dispositions model, with 69 variables preferring the settled dispositions model in all panels and only three variables preferring the active updating model in all panels.

While we cannot account for attitudes and behaviors that are not measured here, these results suggest, consistent with KV, that a wide range of attitudes and behaviors tend to be settled by the time people are old enough to participate in surveys like the GSS. While this is not surprising since we examine the same set of questions as KV, it is important that different statistical assumptions lead to similar substantive conclusions. People might change their responses from wave to wave, but this change does not tend to persist, meaning it reflects temporary changes in opinion or measurement error.[^temp] There is little evidence of even a moderate amount of durable change in opinions in the medium-term (2-4 years).

[^temp]: Within the settled dispositions model, our method cannot distinguish between change that occurs because people lack opinions and are more or less guessing at random [@converse1964], change that occurs because people are influenced by short-term messaging and other considerations [@zaller1992], or changes that occurs because of other forms of measurement error [@ansolabehere2008; @alwin2007]. We assume that these short-term changes are some combination of these forces, which might vary by question.

On specific variables, the approach presented here occasionally suggests different conclusions than those presented in KV, as would be expected under models that make different assumptions, with the model presented here favoring the SDM model in more cases. One reason these differences emerge is because the model presented here uses slightly more data than the KV model. A person who gives the same response in waves 1 and 2 but a different response in wave 3 is mostly useless under the KV model but can factor into the calculation in the model presented here. Because variable-panels can switch from favoring the active updating model to favoring the settled dispositions model with the addition or subtraction of a few cases, this suggests that evidence of the AUM is very weak even when it is favored. In particular, when the AUM is favored, it is because a very small proportion of the population is making durable changes. It is difficult to draw broader conclusions about these discrepancies, as they cut across both question substance and question structure.

While it would be impossible to discuss all 183 variables or 536 variable-panels in depth, there are some broad patterns worth highlighting. Figure \@ref(fig:categories) plots the proportion of variable-panels preferring each model by subject material.[^subjects] KV's results suggest that issues related to civil liberties, gender, and social trust tend to be settled by the time people entered the the GSS. The method outlined above finds similar results, with 86 percent, 89 percent, and 91 percent of variable-panels in these categories, respectively, favoring the settled dispositions models. These are the three categories with the largest proportions of variable-panels favoring the settled dispositions model.

[^subjects]: It should be kept in mind when comparing across subject areas that the questions in subject areas often have different structures, and the questions asked in each category should not be thought of as a random sample of issues in that domain.

```{r categories, fig.cap='Proportion of perferred models, by subject category', fig.align='center'}
load("attitude_vars.Rdata")
winners %>%
  left_join(attitude_vars, by = c("variable"="var")) %>% 
  group_by(small.cat, verdict) %>%
  summarise(count = n()) %>%
  group_by(small.cat) %>%
  mutate(pct = count/sum(count),
         sdm.pct = ifelse(verdict == "SDM", pct, 0)) %>% 
  ggplot(aes(x = reorder(small.cat, sdm.pct), fill = verdict, y = pct)) + 
  geom_bar(stat = "identity", color = "black") + 
  coord_flip() + 
  labs(y = "Proportion of variable-panels", x = "", 
       fill = "Preferred model:") + 
  scale_fill_grey() + 
  theme_minimal() + 
  theme(legend.position = "top")
```

Similarly, KV suggested that public behaviors were more likely to demonstrate active updating, since these behaviors receive social reinforcement that facilitate durable change. In the results presented here, social behaviors such as socializing at a bar and religious behaviors such as church attendance are the two categories that show the highest rates of active updating, with 41 percent and 50 percent of variable-panels in these categories, respectively, showing evidence of active updating.

Another issue worth touching on is variation across panels in favoring the settled dispositions or active updating models within the same question. Of the 176 questions that appear in all three panels, 110 either favor the settled dispositions model or are inconclusive in all three waves (with 69 clearly favoring active updating in all panels). However, 58 variables favor each model in at least one panel.

Some of these discrepancies are likely due to real changes in the social environment that only affect members of one panel. For example, several of the questions with  divergent findings pertain to whether the federal government is spending too little, about right, or too much on various priorities. People's responses might be changing in response to changes in federal spending, changes in who controls the federal government, or changes in public messaging around these issues, which might only happen in certain years.

For questions that demonstrate active updating in at least one panel but have no obvious changing external referent (such as church attendance, religious activity, partisan identification, frequency of prayer, and support for marijuana legalization), durable change might be happening but occur at such low rates that finite samples using imperfect measures do not capture enough respondents undergoing durable change in a particular panel to favor the active updating model. Again, even on generally stable issues we expect some people in the population to be actively updating their behavior -- most people know an adult who has changed their opinion on some issue -- but this should be thought of as the exception.[^recall] Our results suggest that when we observe someone change their response to a survey question, the assumption should be that it will not be a permanent change.

[^recall]: One reason people can recall examples of people in their social networks who have changed beliefs and behaviors is because it is rare and tends to stick out.

```{r, include = FALSE, eval = FALSE}
 winners %>%
   group_by(variable) %>%
  filter(n() > 2) %>%
  # group_by(ds) %>%
  summarise(n_settled = sum(verdict == "SDM"),
            n_active = sum(verdict == "AUM")) %>% filter(n_settled >= 1 & n_active >= 1) %>% 
  group_by(n_settled) %>%
  summarise(n = n())
  ggplot(aes(x = n_settled)) + 
  geom_bar(color = "black", fill = "gray") 
```

This points to a limitation of both this approach and the one used by KV. Neither approach can quantify the proportion of the sample that is undergoing active updating. It is not clear whether, for the variable-panels that show evidence active updating, there is a large or small proportion of respondents making durable changes.

The two approaches share additional limitations. Both approaches assume linearity in the variable being measured. If an outcome is measured on an ordinal scale such as a four-point Likert scale (strongly agree, agree, disagree, strongly disagree), a change from agree to strongly agree is treated the same as a change from disagree to agree, which might or might not be a reasonable assumption. Both approaches also rely on weights to account for differential non-response over the course of the panel. If the mechanism of non-response and active updating are both driven by some unobserved variable, then our results might under-estimate the probability of durable change. That being said, we have no reason to assume that non-response is more likely to be related to active updating than settled dispositions.

The approach presented here and the approach outlined in KV attempt to adjudicate a similar theoretical question: are there more people reporting patterns that look like active updating than we would expect if people behaved according to a settled dispositions model? The difference between the two approaches are sets of assumptions about what counts as an active updating pattern and what counts as a settled dispositions pattern. The fact that different sets of assumptions reach the same substantive conclusion, even under tests very favorable to detecting active updating, strengthens the findings from KV and others [e.g., @vaisey2016] that the settled dispositions model should be thought of as the dominant model for attitude behavior in adults.
